Modal Abstractions for Virtualizing Memory Addresses

### 2023-04-10 background.tex
+

\begin{figure}
    \includegraphics[width=\columnwidth]{pagetables.png}
    \caption{x86-64 page table lookups.}
        \todo[inline]{Colin says: I recommend the cr3 caption say "Physical Base Address" rather than just
                base address.
            If we're saying in the text that we just to L4, L3, L2, L1, then the figure should do the same.
        }
    \label{fig:pagetables}
\end{figure}

@@

Most modal logics use unary operators on propositions to express that a logical claim $P$ is \emph{contingently} true 
in certain other circumstances, such as in other times~\cite{pnueli1977temporal} or places~\cite{gordon2019modal}. That is, for 
different modal operators $M$, $M(P)$ may mean that $P$ is true in another time or place (which may be a physical place 
like another network node~\cite{murphy2008type,gordon2019modal} or a conceptual place like a person's 
mind~\cite{hintikka1962knowledge}). A unifying concept across any modality is that they behave as applicative functors, 
typically satisfying (directly, or as a derived law):
\[ (P\rightarrow Q) \rightarrow M(P) \rightarrow M(Q)\]
\todo[inline]{Ismail, I think this means we have pure intro, $\square P \mathrel{-\ast} [r](\square P)$ if I'm using the right symbol for pure assertions }
\todo[inline,color=red]{The above todo comment isn't quite right.
Purity is about being able to duplicate. You had defined another typeclass/property-of-assertions that meant it didn't
care what address space it was in (like physical pointstos). That's orthogonal to purity.
}
Many modalities, so-called \emph{normal} modalities also possess introduction rules of the form $P\rightarrow M(P)$, 
the classic example being that if $P$ is true, then $P$ is \emph{necessarily} true ($\square P$).

Of particular interest for reasoning about virtual memory are modalities that permit \emph{naming} the alternate 
circumstances, prominently \emph{hybrid} modal logics~\cite{blackburn1995hybrid,areces2001hybrid}, which come equipped 
with assertions of the form $@_\ell(P)$ indicating that $P$ is true in the specific alternate circumstance (Kripke world)
 named by the \emph{nominal} $\ell$.

For our purposes, these are natural candidates to adapt for virtual memory management. We can reinterpret the notion of 
naming an alternate world slightly more loosely, and instead name \emph{address spaces} by the physical address of the 
page table root, since these structures are the physical representations of page tables. Thus in this paper we develop 
the notion that we can represent contingent truth of an assertion via $[r](P)$ indicating that $P$ holds in the address 
space rooted at physical address $r$.
The typical hybrid modality introduction rule, that $P$ and $\ell$ (indicating the current world is $\ell$) imply 
$@_\ell(P)$, has a natural analogue: knowing $P$ and that the current address space is $r$ (i.e., that $r$ is the 
current \texttt{cr3} value) suggests a way to construct $[r](P)$.

>>>>

\begin{figure}
    \includegraphics[width=\columnwidth]{pagetables.png}
    \caption{x86-64 page table lookups.}
    \label{fig:pagetables}
\end{figure}

@@

Most modal logics use unary operators on propositions to express that a logical claim $P$ is \emph{contingently} true 
in certain other circumstances, such as in other times~\cite{pnueli1977temporal} or places~\cite{gordon2019modal}. That is, for 
different modal operators $M$, $M(P)$ may mean that $P$ is true in another time or place (which may be a physical place 
like another network node~\cite{murphy2008type,gordon2019modal} or a conceptual place like a person's 
mind~\cite{hintikka1962knowledge}). A unifying concept across any modality is that they behave as applicative functors, 
typically satisfying (directly, or as a derived law):
\[ (P\rightarrow Q) \rightarrow M(P) \rightarrow M(Q)\]
%\todo[inline]{Ismail, I think this means we have pure intro, $\square P \mathrel{-\ast} [r](\square P)$ if I'm using the right symbol for pure assertions }
%\todo[inline,color=red]{The above todo comment isn't quite right.
%Purity is about being able to duplicate. You had defined another typeclass/property-of-assertions that meant it didn't
%care what address space it was in (like physical pointstos). That's orthogonal to purity.
%}
Many modalities, so-called \emph{normal} modalities also possess introduction rules of the form $P\rightarrow M(P)$, 
the classic example being that if $P$ is true, then $P$ is \emph{necessarily} true with the contingency picked up.%($\square P$).

Of particular interest for reasoning about virtual memory are modalities that permit \emph{naming} the alternate 
circumstances, prominently \emph{hybrid} modal logics~\cite{blackburn1995hybrid,areces2001hybrid}, which come equipped 
with assertions of the form $[\ell](P)$ indicating that $P$ is true in the specific alternate circumstance (Kripke world)
 named by the \emph{nominal} $\ell$.

For our purposes, these are natural candidates to adapt for virtual memory management. We can reinterpret the notion of 
naming an alternate world slightly more loosely, and instead name \emph{address spaces} by the physical address of the 
page table root, since these structures are the physical representations of page tables. Thus in this paper we develop 
the notion that we can represent contingent truth of an assertion via $[r](P)$ indicating that $P$ holds in the address 
space rooted at physical address $r$.
The typical hybrid modality introduction rule, that $P$ and $\ell$ (indicating the current world is $\ell$) imply 
$[\ell](P)$, has a natural analogue: knowing $P$ and that the current address space is $r$ (i.e., that $r$ is the 
current \texttt{cr3} value) suggests a way to construct $[r](P)$. By the choice of \texttt{cr3} identifying the contingency,
we can also identify a fact as a \textit{pure} fact as long as it does not \textit{necessarily} depend on \texttt{cr3}: a truth representing a raw physical memory addressing, unlike virtual-address translation, does not need the \textit{knowledge} of \texttt{cr3}, therefore it can be introduce into the modal context as a pure fact.
